Legion: Expressing Locality and Independence with Logical Regions

### 2012-08-08 doc/sc2012/code_ex.tex
Lots of small edits for the camera ready version of the SC paper

  CircuitPiece pieces[MAX_PIECES];
  for(i = 0; i #$<$# MAX_PIECES; i++) 
    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
  for (t = 0; t #$<$# TIME_STEPS; t++) {
    spawn (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
    spawn (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
    spawn (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
  }
}

void calc_new_currents(CircuitPiece piece):
        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
  foreach(w : piece.rw_pvt)
    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
}

void distribute_charge(CircuitPiece piece, float dt):
        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
  foreach(w : piece.rw_pvt) {
    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
  }
}

>>>>

  CircuitPiece pieces[MAX_PIECES];
  for(i = 0; i #$<$# MAX_PIECES; i++) 
    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
  for (t = 0; t #$<$# TIME_STEPS; t++) {
    spawn (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
    spawn (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
    spawn (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
  }
}
                           // ROE = Read-Only-Exclusive
void calc_new_currents(CircuitPiece piece):
        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
  foreach(w : piece.rw_pvt)
    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
}
                          // RdA = Reduce-Atomic
void distribute_charge(CircuitPiece piece, float dt):
        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
  foreach(w : piece.rw_pvt) {
    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
  }
}

### 2012-08-08 doc/sc2012/conclusion.tex
Lots of small edits for the camera ready version of the SC paper

We have presented Legion, a programming model for expressing
the abstract locality and independence properties of program
data by using logical regions.  The Legion programming model enabled us to implement
a distributed Software-Out-of-Order Processor (SOOP) scheduling 
algorithm.  The SOOP leveraged the locality and independence properties
capture by logical regions to efficiently execute programs.  We also
described a novel mapping interface that guides the SOOP execution
and allows tuning of program performance independent of correctness.
We demonstrated that our SOOP runtime implementation
of Legion was faster than existing implementations of
three benchmarks and that it was able to scale to distributed, heterogeneous
machines.

>>>>

We have presented Legion, a programming model for expressing
the abstract locality and independence properties of program
data by using logical regions.  The Legion programming model enables 
a distributed Software-Out-of-Order Processor (SOOP) scheduling 
algorithm.  The SOOP algorithm leverages the locality and independence properties
captured by logical regions to efficiently execute programs.  We also
described a novel mapping interface that guides the SOOP execution
and allows tuning of program performance independent of correctness.
We demonstrated that our SOOP runtime implementation
of Legion was faster than existing implementations of
three benchmarks and that it was able to scale on distributed, heterogeneous
machines.

### 2012-08-08 doc/sc2012/experiments.tex
Lots of small edits for the camera ready version of the SC paper

Figure~\ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
to a hand-coded single-GPU implementation written in CUDA.  The hand-coded implementation is
able to keep the entire simulation state in fast framebuffer memory.  Each line shows the scaling of
a particular problem size as the number of nodes is varied.  Our results demonstrate
excellent strong scaling, with speedups of 39.0X for the small problem on 48 GPUs and 
62.5X for the larger problem size on 96 GPUs.  The inset in the graph shows the relative
performance for small numbers of GPUs.  On a single GPU, our Legion implementation
is within 5\% of the performance of the hand-coded simulation.

Figure~\ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
all nodes) spent in the application kernels compared to the various pieces of the Legion
SOOP runtime.  As the node count increases, the non-communication overhead stays relatively constant.
As expected the communication overhead grow linearly with number of nodes.

@@

Achieving high-performance on this application is particularly
challenging for several reasons.  First, the application has a very
high communication-to-computation ratio which, for a fixed problem
size, begins as being memory bound and with increasing node count
becomes network bound as the perimeter-to-area ratio of cell grids
increases.  Second, when choosing how to partition cells into grids,
the programmer must consider the locality between cells within a
level as well as across levels.  For
cross-level cell dependences, optimal mapping decisions can only be made at
runtime as the location of refinements are dynamically determined.
Finally, this application has parallelism both
between tasks running at the same level and tasks running across
levels, leading to complicated input-dependent data dependences.

>>>>

Figure~\ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
to a hand-coded single-GPU implementation written in CUDA.  The hand-coded implementation is
able to keep the entire simulation state in fast framebuffer memory.  Each line shows the scaling of
a particular problem size as the number of compute nodes is varied.  Our results demonstrate
excellent strong scaling, with speedups of 39.0X for the small problem on 48 GPUs and 
62.5X for the larger problem size on 96 GPUs.  The inset in the graph shows the relative
performance for small numbers of GPUs.  On a single GPU, our Legion implementation
is within 5\% of the performance of the hand-coded simulation.

Figure~\ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
all nodes) spent in the application kernels compared to the various pieces of the Legion
SOOP runtime.  As the node count increases, the non-communication overhead stays relatively constant.
As expected, the communication overhead grows quadratically with the number of nodes.

@@

Achieving high-performance on this application is particularly
challenging for several reasons.  First, the application has a very
high communication-to-computation ratio which, for a fixed problem
size, begins as being memory bound and, with increasing node count,
becomes network bound as the perimeter-to-area ratio of cell grids
increases.  Second, when choosing how to partition cells into grids,
the programmer must consider the locality between cells within a
level as well as across levels.  For
cross-level cell dependences, optimal mapping decisions can only be made at
runtime as the location of refinements are dynamically determined.
Finally, this application has parallelism both
between tasks running at the same level and tasks running across
levels, leading to complicated input-dependent data dependences.

### 2012-08-08 doc/sc2012/legion.tex
Lots of small edits for the camera ready version of the SC paper

\item We present results of experiments on three applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  heat diffusion equation and a circuit simulation.
  We compare each application with the best reference
  versions on three different clusters of multicore processors with
  GPUs, including the Keeneland supercomputer \cite{Keeneland}.
%
% FIXME Give the same numbers as in the abstract.
%

>>>>

\item We present results of experiments on three applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  heat diffusion equation, and a circuit simulation.
  We compare each application with the best reference
  versions on three different clusters of multicore processors with
  GPUs, including the Keeneland supercomputer \cite{Keeneland}.
%
% FIXME Give the same numbers as in the abstract.
%

### 2012-08-08 doc/sc2012/related.tex
Lots of small edits for the camera ready version of the SC paper

@@

X10 is another parallel programming language designed to operate on
distributed memory machines \cite{X1005}.  X10's {\em places} enable
programmers to talk about where to place both data and tasks.
However, once data and tasks have been placed they are fixed, which
mandates that data movement be explicitly managed by user level code
or implicitly by the compiler or runtime \cite{X1008}.  Recently X10
has introduced regions into the compiler's
intermediate representation \cite{X1011}.  Unlike Legion, regions in
X10 are not visible to the programmer but are inferred from high level
arrays through static analysis.  X10 provides support for clusters of
GPUs \cite{X10GPU}, but requires the programmer to write all code
managing data movement through both the cluster and GPU memory
hierarchies.

Deterministic Parallel Java (DPJ) is a parallel extension of Java
that, like Legion, uses regions to express locality and does
dependence analysis at the granularity of conflicts on region
arguments to functions \cite{Bocchino09}.  As the name suggests, one
of the primary goals of DPJ is to provide a programming model that
guarantees determinism while also supporting parallelism.
As a result the programming model is more restrictive than Legion, as
there is no notion of weaker coherence properties of regions than
exclusive access (which is essential to determinism).  DPJ can express
the populate-then-partition style of using regions, but not the
partition-then-populate; overall, the language is more static, and
thus provides more guarantees and less flexibility than Legion.  
DPJ efforts have so far focused on JVM implementations on shared memory machines.
There have also been several other recent efforts that propose task schedulers
based on dynamic detection of memory conflicts, but these also have assumed
underlying shared memory hardware \cite{VTN11,Tzenakis12,EYJD12}.

>>>>

The SSMP programming model is the most similar work to Legion \cite{Perez10}.  
Like Legion, SSMP supports dynamic detection of dependences between tasks based 
on data requirements.  However, SSMP only supports a single disjoint rectilinear 
partition of an array, unlike Legion which supports multiple
arbitrary partitions of regions.  Furthermore, the SSMP runtime must perform dependence
checks between every pair of tasks created in the system.  Legion's programming model only
requires dependence checks between tasks with the same parent task which enables scalable nested
parallelism in distributed machines.  SSMP only operates on shared memory machines.

@@

X10 is another parallel programming language designed to operate on
distributed memory machines \cite{X1005}.  X10's {\em places} enable
programmers to talk about where to place both data and tasks.
However, once data and tasks have been placed they are fixed, which
mandates that data movement be explicitly managed by user level code
or implicitly by the compiler \cite{X1008}.  Recently X10
has introduced regions into the compiler's
intermediate representation \cite{X1011}.  Unlike Legion, regions in
X10 are not visible to the programmer but are inferred from high level
arrays through static analysis.  X10 provides support for clusters of
GPUs \cite{X10GPU}, but requires the programmer to write all code
managing data movement through both the cluster and GPU memory
hierarchies.

Deterministic Parallel Java (DPJ) is a parallel extension of Java
that, like Legion, uses regions to express locality and does
dependence analysis at the granularity of conflicts on region
arguments to functions \cite{Bocchino09}.  As the name suggests, one
of the primary goals of DPJ is to provide a programming model that
guarantees determinism while also supporting parallelism.
As a result the programming model is more restrictive than Legion, as
there is no notion of weaker coherence properties of regions than
exclusive access (which is essential to determinism).  DPJ can express
the populate-then-partition style of using regions, but not the
partition-then-populate; overall, the language is more static, and
thus provides more guarantees and less flexibility than Legion.  
DPJ efforts have so far focused on JVM implementations on shared memory machines.

There have also been several other efforts that use partitions to either avoid
or detect conflicts in shared memory programs dynamically.
Object assemblies are a mechanism for partitioning a shared memory
heap to enable parallel execution \cite{Lublinerman09}.  Object assemblies
only support a disjoint partitioning of the heap unlike Legion which
allows for multiple partitions of any data structure.  Legion further uses 
the information from partitions to operate on distributed architectures.

Galois is programming system based on optimistic parallelism that dynamically detects 
memory conflicts between concurrent threads at the very
fine granularity of object accesses.  To achieve reasonable performance
Galois supports a partitioning interface for breaking up data structures \cite{Kulkarni08}.  
Galois relies on shared memory
and only leverages partitioning for coarser locking and conflict detection.
Several other recent efforts have proposed task schedulers
based on dynamic detection of memory conflicts, but these also have assumed
underlying shared memory hardware \cite{VTN11,Tzenakis12,EYJD12}.

### 2012-08-08 doc/sc2012/soop.tex
Spelling corrections
Lots of small edits for the camera ready version of the SC paper

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions:
just as an instruction scheduler is constrained by register dependences,
the SOOP is constrained by region dependences.
The SOOP  is pipelined, distributed and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks, which is non-trivial
because the same data may be in multiple different regions or subregions.

@@

Detecting mapping dependences between a newly registered task $t$ and
a previously registered task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region used by $t'$ that
may {\em alias} (may share data with) a logical region used by $t$, the privileges and
coherence modes are compared to determine whether a dependence exists.
If both regions need only read privileges there is never a dependence,
but if either task needs write or reduction privileges, the coherence
modes are compared using the table in Figure~\ref{fig:dependence}.
{\tt Dep} indicates dependence while {\tt None}
indicates independence.  {\tt Same} is a dependence unless the two tasks
use the same physical instance of the logical region.  Since tasks
haven't mapped physical instances at this point in the pipeline,
{\tt Same} is always a mapping dependence.
%\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
{\tt Cont} 
indicates a dependence contingent on the privileges of the two tasks
(e.g. an Atomic Read-Write task and Simultaneous Read-Only task will not have a dependence).

@@

There are two steps to mapping a region $r$ used by
task $t$.  First, the mapper must ensure there is at least one {\em valid}
physical instance of $r$ for $t$ to use.  Not all intances have up-to-date
data at all times.  For example, if an instance of a subregion of $r$ has been written by a previous task,
it is necessary to copy that subregion back to an instance of $r$ so that $t$
sees the correct data. In the second step, the mapping interface selects one of the
valid physical instances of $r$ for $t$ to use or creates a new one.  
We focus on the first step; the mapping interface is discussed further in Section~\ref{sec:mapping}.

>>>>

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions:
just as an instruction scheduler is constrained by register dependences,
the SOOP is constrained by region dependences.
The SOOP  is pipelined, distributed, and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks, which is non-trivial
because the same data may be in multiple different regions or subregions.

@@

Detecting mapping dependences between a newly registered task $t$ and
a previously registered task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region used by $t'$ that
may {\em alias} (may share data with) a logical region used by $t$, the privileges and
coherence modes are compared to determine whether a dependence exists.
If both regions need only read privileges there is never a dependence,
but if either task needs write or reduction privileges, the coherence
modes are compared using the table in Figure~\ref{fig:dependence}.
{\tt Dep} indicates dependence while {\tt None}
indicates independence.  {\tt Same} is a dependence unless the two tasks
use the same physical instance of the logical region.  Since tasks
have not mapped physical instances at this point in the pipeline,
{\tt Same} is always a mapping dependence.
%\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
{\tt Cont} 
indicates a dependence contingent on the privileges of the two tasks
(e.g. an Atomic Read-Write task and Simultaneous Read-Only task will not have a dependence).

@@

There are two steps to mapping a region $r$ used by
task $t$.  First, the mapper must ensure there is at least one {\em valid}
physical instance of $r$ for $t$ to use.  Not all instances have up-to-date
data at all times.  For example, if an instance of a subregion of $r$ has been written by a previous task,
it is necessary to copy that subregion back to an instance of $r$ so that $t$
sees the correct data. In the second step, the mapping interface selects one of the
valid physical instances of $r$ for $t$ to use or creates a new one.  
We focus on the first step; the mapping interface is discussed further in Section~\ref{sec:mapping}.
